---
title: "Illustrating the hazard function"
author: "Steve Simon"
date: "Friday, July 15, 2016"
output: html_document
---

## Abstract

When you have data measuring the time to an event, you can examine the relationship between various predictor variables and the time to the event using a Cox proportional hazards model. In this talk, you will see what a hazard function is and describe the interpretations of increasing, decreasing, and constant hazard. Then you will examine the log rank test, a simple test closely tied to the Kaplan-Meier curve, and the Cox proportional hazards model.

## What is the hazard function?

```{r weibull_density}
age <- seq(1, 100, by=1)
age <- round(age, 1)
density <- dweibull(age, 2, 35)
plot(age, density, type="l")
```

Here is a probability density function that might represent the
survival probability for an individual. It is simple from a 
mathematical perspective, but may not reflect actual survival
probabilities for a realistic individual.

In a density function, probabilities are computed by measuring
the area under the curve. So the probability that you will die
as a young adult (between ages 20 and 30) is

```{r area_1}
pct <- function(x) (paste(round(100*x), "%", sep=""))
plot(age, density, type="l")
age1 <- 20
age2 <- 30
age_mean <- (age1+age2) / 2
segments(age1, 0, age1, density[age==age1])
segments(age2, 0, age2, density[age==age2])
age_range <- (age>=age1) & (age<=age2)
polygon(x=c(age2, age1, age[age_range]), y=c(0, 0, density[age_range]), border=NA, col="gray")
p1 <- pweibull(age1, 2, 35)
p2 <- pweibull(age2, 2, 35)
probability <- p2 - p1
text(age_mean, 0.5*density[age==age_mean], pct(probability))
```

The probability that you will die as a middle aged adult 
(between ages 40 and 70) is

```{r area_2}
plot(age, density, type="l")
age3 <- 40 
age4 <- 70
age_mean <- (age3+age4) / 2
segments(age3, 0, age3, density[age==age3])
segments(age4, 0, age4, density[age==age4])
age_range <- (age>=age3) & (age<=age4)
polygon(x=c(age4, age3, age[age_range]), y=c(0, 0, density[age_range]), border=NA, col="gray")
p3 <- pweibull(age3, 2, 35)
p4 <- pweibull(age4, 2, 35)
probability <- p4 - p3
text(age_mean, 0.5*density[age==age_mean], pct(probability))
```

Now these are interesting probabilities, but they are not directly comparable for two reasons.

1. The probabilities are over different time frames.

2. The probabilities should be adjusted to account for the smaller number of survivors at the older ages.

To fix the first problem, you need to look at the rate of death. This is the number of deaths
per million people divided by the number of days in the time interval. The choice of a 
million people instead of, say 10 thousand, and the choice of days instead of say, months
or years, is a bit arbitrary. But as long as you are consistent, the comparisons are fairer.

To fix the second problem, you need to adjust the rate by the number of people who have survived
up to age 18, 40, or 70.

```{r basic calculations, echo=FALSE}
f1 <- round(10000*(p2-p1))
d1 <- 365*(age2-age1) + (age2-age1) %/% 4
r1 <- round(f1/d1, 2)
x1 <- round(10000*p1)
y1 <- 10000 - x1
h1 <- round(r1 * 10000 / y1, 2)

f3 <- round(10000*(p4-p3))
d3 <- 365*(age4-age3) + (age4-age3) %/% 4
r3 <- round(f3/d3, 2)
x3 <- round(10000*p3)
y3 <- 10000 - x3
h3 <- round(r3 * 10000 / y3, 2)
```

If you have 10,000 people, you'd see `r f1` funerals between the ages of 
`r age1` and `r age2` and since there are `r d1` days in this interval,
you'd have a rate of `r r1` (`r f1` / `r d1`) funerals per day. But
of ten thousand people, there would be `r x1` who died before age `r age1` and
`r y1` survivors. So the rate among 10,000 people alive at age `r age1` is 
`r h1` (`r r1`*10000 / `r y1`) funerals per day.
 
Middle aged adults (`r age3`-`r age4`) have about the same probability, but
it's over a longer period of time. Counterbalancing this, there are not quite
as many people who survive to 40. Here's the math.
 
For 10,000 people, you'd see `r f3` funerals over a span of `r d3` days. That's
a rate of `r r3` (`r f3` / `r d3`) funerals per day. There were `r x3` deaths
before age `r age3` and `r y3` survivors. So the rate among 10,000 people alive
at age `r age3` is `r h3` (`r r3`*10000 / `r y3`) funerals per day.

The hazard rate is defined in a similar way. 

h(t) = (P[t $\le$ T $\le$ T+$\Delta$t] / $\Delta$t) / P[T $\ge$ t]

where $\Delta$t is very small.

It's a rate since it involves a measure of time, so the hazard rate does not
necessarily have to be bounded between 0 and 1 like a probability would be.

The hazard rate is mathematically equivalent to the density function divided 
by the survival function. You need some basic calculus to establish this

```{r trio}
plot(age, density, type="l")
title("Density")
survival_probability <- 1 - pweibull(age, 2, 35)
plot(age, survival_probability, type="l")
title("Survival")
plot(age, density / survival_probability, type="l")
title("Hazard rate")
```

This is an example of an increasing hazard rate. Your death rate is low at any 
particular time point, but the rate increases over time. People
who talk about survival with respect to machines refer to this 
as "new better than used."

There are examples of a decreasing hazard rate or "used better than new." 

```{r decreasing hazard}
density <- dweibull(age, 0.5, 140)
plot(age, density, type="l")
title("Density")
survival_probability <- 1 - pweibull(age, 0.5, 140)
plot(age, survival_probability, type="l")
title("Survival")
plot(age, density / survival_probability, type="l")
title("Hazard rate")
```

This is reminiscent of the Nietzsche quote "That which does not kill us makes
us stronger." This does not mean that you'll never die, but your short term risk
of dying is greatest when you are young and small. From the engineering perspective
if a machine has a decreasing haard rate, that means that it gets battle
hardened as it get older and becomes more reliable over time.

```{r constant hazard}
density <- dexp(age, rate=1/70)
plot(age, density, type="l")
title("Density")
survival_probability <- 1 - pexp(age, rate=1/70)
plot(age, survival_probability, type="l")
title("Survival")
plot(age, density / survival_probability, type="l")
title("Hazard rate")
```

The exponential distribution is said to have a "memoryless" property. What this
means is that the short term risk of death is unrelated to age. In other words a
new machine performs no better or no worse than a used machine.

Another common pattern for the hazard rate is called the bathtub hazard. This is a
hazard that starts out high early, decreases, flattens out for a while and then
gradually increases later in life. The bathtub hazard is the one that most closely
resembles the human experience.

You can find a nice illustration of the bathtub hazrd at

https://commons.wikimedia.org/wiki/File:Bathtub_curve.svg

## Why is the hazard function important?

1. Knowledge of the hazard function for an industrial product can help you decide on how and when to employ preventive maintenance. 

2. It can also help you extrapolate survival patterns beyond the range of observed values, such as for long-term reliability testing.

3. If you are planning a new research study, the hazard function can help you decide how many patients to study and the length of follow-up for these patients.

4. Assumptions about the hazard function are critical for regression models of survival.

## The heroin data set.

On the Australian Data and Story Library website, you can find a data set on the amount of time that a patient spens in a rehabilitation clinic. You can read a description of the data at

--> http://www.statsci.org/data/oz/heroin.html

and the tab delimited data is at

--> http://www.statsci.org/data/oz/heroin.txt

Note that for this study, a small time-to-event is good.

```{r read}
me_raw <- read.table("heroin.txt", header=TRUE)
head(me_raw)
m1 <- me_raw[ , 1:6]
m2 <- me_raw[ , 7:12]
names(m2) <- names(m1)
me <- rbind(m1, m2)
head(me)
tail(me)
dim(me)
summary(me$Time)
summary(me$Dose)
table(me$Dose)
table(me$Clinic)
table(me$Status)
table(me$Prison.)
```

In R, the default is that 0 represents a censored value and 1 represents an event.

Before you fit a Cox proportional hazards model, you should look at the Kaplan-Meier curve and the logrank test.

```{r survival}
library("survival")
su <- Surv(me$Time, me$Status)
print(su)
plot(survfit(su~1))
quantile(survfit(su~1))

plot(survfit(su~me$Clinic))
quantile(survfit(su~me$Clinic))
survdiff(su~me$Clinic)

plot(survfit(su~me$Prison.))
quantile(survfit(su~me$Prison.))
survdiff(su~me$Prison.)

dose_quartiles <- cut(me$Dose, c(0, 50, 60, 70, 999))
plot(survfit(su~dose_quartiles))
quantile(survfit(su~dose_quartiles))
survdiff(su~dose_quartiles)
```

Here's how you calculate the log rank test

```{r hand-calculation}
library("tibble")
tb <- table(me$Time[me$Status==1], me$Prison.[me$Status==1])
tim <- as.numeric(dimnames(tb)[[1]])
o1j <- tb[, 1]
o2j <- tb[, 2]
oj <- o1j+o2j
n1j <- rep(NA, length(tim))
n2j <- rep(NA, length(tim))
for (j in 1:length(tim)) {
  n1j[j] <- sum(me$Time[me$Prison.==0]>=tim[j])
  n2j[j] <- sum(me$Time[me$Prison.==1]>=tim[j])
}
nj <- n1j+n2j
pj <- round(n1j/nj, 2)
e1j <- round(oj*pj, 2)
v1j <- round(oj*pj*(1-pj) * (nj-oj)/(nj-1), 2)
tail(tibble(tim, o1j, o2j, oj, n1j, n2j, nj, pj, e1j, v1j))
logrank_z <- sum(o1j-e1j) / sqrt(sum(v1j))
logrank_x2 <- logrank_z^2
```

The Cox proportional hazards model allows a lot more flexibility than the logrank test. You can include a continuous predictor variable (no need to create quartiles for dose) and multiple predictors (allowing for a risk adjusted model, for example).

```{r cox}
coxph(su~me$Prison.)
coxph(su~dose_quartiles)
coxph(su~me$Dose)
coxph(su~strata(me$Clinic)+me$Prison.+me$Dose)
```

Save everything for later re-use

```{r save-everything}
save.image("illustrate_hazard.RData")
```

