---
title: "Illustrating the hazard function"
author: "Steve Simon"
date: "Friday, July 15, 2016"
output: html_document
---

This program shows a simple example of the hazard function.n R code chunk like this:

```{r weibull_density}
age <- seq(1, 100, by=1)
age <- round(age, 1)
density <- dweibull(age, 2, 35)
plot(age, density, type="l")
```

Here is a probability density function that might represent the
survival probability for an individual. It is simple from a 
mathematical perspective, but may not reflect actual survival
probabilities for a realistic individual.

In a density function, probabilities are computed by measuring
the area under the curve. So the probability that you will die
between ages 20 and 30 is

```{r area_1}
pct <- function(x) (paste(round(100*x), "%", sep=""))
plot(age, density, type="l")
age1 <- 20
age2 <- 30
age_mean <- (age1+age2) / 2
segments(age1, 0, age1, density[age==age1])
segments(age2, 0, age2, density[age==age2])
age_range <- (age>=age1) & (age<=age2)
polygon(x=c(age2, age1, age[age_range]), y=c(0, 0, density[age_range]), border=NA, col="gray")
probability <- pweibull(age2, 2, 35) - pweibull(age1, 2, 35)
text(age_mean, 0.5*density[age==age_mean], pct(probability))
```

The probability that you will die between ages 40 and 60 is

```{r area_2}
plot(age, density, type="l")
age1 <- 40
age2 <- 60
age_mean <- (age1+age2) / 2
segments(age1, 0, age1, density[age==age1])
segments(age2, 0, age2, density[age==age2])
age_range <- (age>=age1) & (age<=age2)
polygon(x=c(age2, age1, age[age_range]), y=c(0, 0, density[age_range]), border=NA, col="gray")
probability <- pweibull(age2, 2, 35) - pweibull(age1, 2, 35)
text(age_mean, 0.5*density[age==age_mean], pct(probability))
```

and the probability that you will die after age 80 is 

```{r area_3}
plot(age, density, type="l")
age1 <- 80
age2 <- 100
age_mean <- (age1+age2) / 2
segments(age1, 0, age1, density[age==age1])
segments(age2, 0, age2, density[age==age2])
age_range <- (age>=age1) & (age<=age2)
polygon(x=c(age2, age1, age[age_range]), y=c(0, 0, density[age_range]), border=NA, col="gray")
probability <- pweibull(age2, 2, 35) - pweibull(age1, 2, 35)
text(age_mean, 10*density[age==age_mean], pct(probability))
```

```{r bars}
yearly_probability <- pweibull(age, 2, 35)-c(0,pweibull(age[-100], 2, 35))
sum(yearly_probability)
plot(age,yearly_probability,type="s")
segments(age, 0, age, yearly_probability)
```
